\Large{Adversarial Variational Optimization of Non-Differentiable Simulators}

### 2017-07-12 tex/paper.tex
stopping for dinner.

Algorithm~\ref{alg:avo} represents the simplest version of AVO; however, the variance of the noisy estimator of the gradients may be too large to be useful in many problems. Variance reduction techniques such as Rao-Blackwellization and Control Variates can be applied here as they are in black box variational inference~\citep{ranganath2014black}.

@@

This work sits at the intersection of several lines of work related to 
likelihood-free inference, approximate Bayesian computation (ABC),
implicit generative networks, and variational inference.  Our focus
is on extending existing work to non-differentiable generative models
that do not admit a tractable density.

Viewed from the literature around implicit generative models based on neural networks, 
the proposed method can be considered as a direct adaptation of
generative adversarial networks~\citep{goodfellow2014generative,2017arXiv170104862A, 2017arXiv170107875A} to
non-differentiable simulators using variational optimization~\cite{2012arXiv1212.4507S,staines2013optimization}.

From the point of view of likelihood-free inference, the closest to our work are procedures that rely on a classifier to estimate the
discrepancy between the observed data and the model distributions. For example,
\citep{gutmann2012noise} uses non linear logistic regression for fitting
unnormalized differentiable statistical models. In the likelihood-free
setup, \citep{cranmer2015approximating,cranmer2016experiments,2016arXiv161110242D} estimate likelihood
ratios through supervised classification, which can in turn be used for
parameter inference in combination with a gradient-free optimization algorithm.
Similarly, \citep{gutmann2017likelihood} makes use of classification accuracy as
a summary statistics for approximate Bayesian computation.

@@

There have been efforts to extend variational inference to intractable likelihoods; however, they require restrictive assumptions. 
In \citep{tran2017variational}, the authors consider Variational Bayes with an intractable likelihood. In that approach "the only requirement is that the intractable likelihood can be estimated unbiasedly.'' In the case of many simulators it is not clear how one would obtain the unbiased estimator $\hat{p}(\bfx|\bftheta)$. In \citep{2017arXiv170208235H} the author defines implicit distributions to be ``probability models whose probability density function may be intractable, but there is a way to 1) sample from them exactly and/or calculate and approximate expectations under them, and 2) calculate or estimate gradients of such expectations with respect to model parameters.'' This notion of implicit model does not include most scientific simulators as 2) is not satisfied. Furthermore, the  \texttt{JC-Adv} and \texttt{JC-Den} algorithm defined there applies to implicit models, but, as the author points out, is aimed at variational inference and ``does not provide a direct way to learn model parameters $\bftheta$''. Lastly, in \citep{meeds2015hamiltonian}, the authors introduce Hamiltonian ABC which applies in the likelihood-free setting, and they estimate gradients

>>>>

Algorithm~\ref{alg:avo} represents the simplest version of AVO; however, the variance of the noisy estimator of the gradients may be too large to be useful in many problems. Variance reduction techniques such as Rao-Blackwellization, Control Variates, and quasi-Monte Carlo can be applied here as they are in black box variational inference~\citep{ranganath2014black}.

@@

This work sits at the intersection of several lines of research related to 
likelihood-free inference, approximate Bayesian computation (ABC),
implicit generative networks, and variational inference.  
%Our focus
%is on extending existing work to non-differentiable generative models
%that do not admit a tractable density.
Viewed from the literature around implicit generative models based on neural networks, 
the proposed method can be considered as a direct adaptation of
generative adversarial networks~\citep{goodfellow2014generative,2017arXiv170104862A, 2017arXiv170107875A} to
non-differentiable simulators using variational optimization~\cite{2012arXiv1212.4507S,staines2013optimization}.

ALI~\citep{dumoulin2016adversarially}, BiGANs ~\citep{donahue2016adversarial}, Adversarial Variational Bayes (AVB)~\citep{DBLP:journals/corr/MeschederNG17}, and, equivalently, the \texttt{PC-Adv} algorithm of  \citep{2017arXiv170208235H})  are a recent extension to GANs that add an inference network to the generative model.  The BiGAN, AVB, and PC-Adv setups are quite similar, but those models assume a tractable density $p(\bfx|\bftheta)$ that is differentiable with respect to $\bftheta$, which are not satisfied in the likelihood-free setting. Our lifting of the non-differentiable simulator $p(\bfx|\bftheta)$ to the variational program $q(\bfx | \bftheta)$ provides the ability to differentiate expectations with respect to $\bfpsi$ as in Eqn~\ref{eqn:approx-grad}; however, the density $q(\bfx | \bftheta)$ is still intractable. Moreover, the latent space $\mathcal{Z}$ of many real-world simulators is complicated and not amenable to neural a recognition model. One similarity to AVB is that the ELBO is rewritten such that the intractable term $\log p(\bfz) ? \log q_\bfphi(\bfz | \bfx)$ is replaced with an optimized neural network $T(x,z)$, which structurally similar to the role of the critic $d$ in the Kantorovich-Rubinstein dual description of the WGAN.

From the point of view of likelihood-free inference, where intractable non-differentiable simulators are the norm, our contributions are four fold. First is the process of lifting expectations with respect to the non-differentiable simulator $\mathbb{E}_{\tilde{\bfx} \sim p(\bfx | \bftheta)}$ to a differentiable expectations with respect to the variational program $\mathbb{E}_{\tilde{\bfx} \sim q(\bfx | \bfpsi)}$. Secondly, is the introduction of a form of variational inference that works in a likelihood-free setting. Thirdly, the AVI-MC algorithm can be viewed as a form of empirical Bayes where the prior is optimized based on the data. Fourthly, the AVL algorithm is not a Bayesian inference algorithm, but instead provides point estimates $\bftheta^*$.

Perhaps the closest to our work is~\cite{bernton2017inference}, which uses Wasserstein distance for a rejection sampler in an ABC-like setup (as opposed to variational inference).  They emphasize that this approach allows one to eliminate the summary statistics typically used in ABC and calculate the Wasserstein distance explicitly without use Kantorovich-Rubinstein duality and the critic $d$. For high-dimensional data, they note that this is computationally expensive and introduce an approximation based on projection onto Hilbert space filling curves. Their Remark 5.1 points out that~\citep{montavon2016wasserstein} proposed an approximation of the gradient of an entropy-regularized Wasserstein distance, which uses a similar duality. They note that ``Unfortunately, it is not applicable in the setting of purely generative models, as it involves point-wise evaluations of the derivative of the log-likelihood.'' \kcnote{Can we evaluate their Eq. 6 with our Eq. 9} Thus, our contribution is to provide gradients of an approximate MEWE by taking expectations with the variational program $q(\bfx | \bf\theta)$. This paired with the critic Kantorovich-Rubinstein dual formulation of the Wasserstein distance using the critic $d$ allows us to work in high dimensions without summary statistics and optimize $\bfpsi$ with stochastic gradient descent.

\kcnote{Could we actually replace the optimization of the critic with a direct Wasserstein distance calculation in fewer dimensions. If the data is drawn from the variational program, it would still be differentiable with respect to $\bfpsi$}

Within the literature on ABC and likelihood-free inference, 
the closest to our work are procedures that rely on a classifier to estimate the
discrepancy between the observed data and the model distributions. For example,
\citep{gutmann2012noise} uses non linear logistic regression for fitting
unnormalized differentiable statistical models. Similarly, \citep{cranmer2015approximating,cranmer2016experiments,2016arXiv161110242D} estimate likelihood
ratios through supervised classification, which can in turn be used for
 Bayesian or frequentist inference. Similarly, \citep{gutmann2017likelihood} 
 makes use of classification accuracy as
a summary statistics for approximate Bayesian computation.

@@

There have been efforts to extend variational inference to intractable likelihoods; however, they require restrictive assumptions. 
In \citep{tran2017variational}, the authors consider Variational Bayes with an intractable likelihood. In that approach "the only requirement is that the intractable likelihood can be estimated unbiasedly.'' In the case of many simulators it is not clear how one would obtain the unbiased estimator $\hat{p}(\bfx|\bftheta)$.
\kcnote{They suggest to use ABC likelihood... is this only unbiased as $\epsilon \to 0$?}

 In \citep{2017arXiv170208235H} the author defines implicit distributions to be ``probability models whose probability density function may be intractable, but there is a way to 1) sample from them exactly and/or calculate and approximate expectations under them, and 2) calculate or estimate gradients of such expectations with respect to model parameters.'' This notion of implicit model does not include most scientific simulators as 2) is not satisfied. Furthermore, the  \texttt{JC-Adv} and \texttt{JC-Den} algorithm defined there applies to implicit models, but, as the author points out, is aimed at variational inference and ``does not provide a direct way to learn model parameters $\bftheta$''. Lastly, in \citep{meeds2015hamiltonian}, the authors introduce Hamiltonian ABC which applies in the likelihood-free setting, and they estimate gradients
